The effects of host galaxy properties on merging compact binaries detectable by LIGO

### 2016-07-13 paper/paper.tex
paper: minor changes to intro to better suit new scope

% POINT: Huge detail possible
The host galaxies of distant short GRBs have already been extensively investigated, with the associations being used to
draw preliminary conclusions about their progenitors  \citep{2014ARAA..52...43B}.   
%Of necessity, these analyses  on bulk correlations, often stacked to increase significance across multiple sources
%\editremark{clean up/don't piss off   Edo}.  
Unlike GRBs, detected gravitational wave sources will be limited by the range of LIGO to the local universe; for
example,  binary neutron star sources should be closer than $400\unit{Mpc}$.  Due to their proximity, each host galaxy
can be explored at great depth and detail via  position-resolved spectroscopy, enabling detailed position-resolved star-formation
and chemical evolution histories  \citep[see,e.g.][]{2009MNRAS.396..462K,2014MNRAS.444..336C,
CALIFA,CALIFA2}.  
% POINT: But is it useful?
In this work, we demonstrate by concrete example that detailed analysis of individual galaxies can be essential to
investigate key physical questions about the origin of compact binary mergers.

This paper is organized as follows. 
In \S \ref{sec:sims}, we describe detailed hydrodynamical simulations of  several galaxies, including two of Milky Way-mass and two dwarfs.  Though the two Milky Way-like galaxies are morphologically similar at $z = 0$, we
show that their star formation and chemical evolution history have subtle but important differences due to their
distinctive merger histories.   To demonstrate
these differences have a practical impact,  in \S \ref{sec:model}, we introduce a simple,
metallicity-dependent  phenomenological model to calculate the present-day rate and mass distribution of compact binary
mergers from  a galaxy's known history.   In \S \ref{sec:results:BBH}, we use this model to investigate  the compact
binary  coalescence rate
dependence on the galaxy's assembly history.   
%% We also demonstrate that even the coalescence rates of binary neutron stars (NSNS) and black hole-neutron
%% star (BH-NS) binaries also depend surprisingly sensitively on the assembly history of each galaxy. 
In \S~\ref{sec:Discussion} we discuss the implications of our study for the interpretation of host galaxy associations
identified via transient multmiessenger astronomy, in the short and long term.  We summarize our results in \S~ \ref{sec:conclude}.
%% In \S \ref{sec:results} we show that while these two galaxy formation models should produce comparable detectable binary neutron star
%% populations, these two scenarios could produce dramatically different populations of electromagnetically detectable
%% merging black hole-neutron star binaries. 
%

@@

  The simulation h258, on the other hand, has a much
  more active merger history.  At $z = 1$ there is a 1:1 merger event;
  a gas disk rapidly reforms following the collision (see
  \citet{Governato09}), resulting in a massive disk galaxy at $z = 0$
  which looks remarkably similar to the Milky Way and to the other
  simulation, h277.  Prior to the $z = 1$ merger, each of the two
  progenitor galaxies actually experiences its own additional major
  merger events around $z = 3$.  The combination of
  the series of major mergers, plus a number of minor interactions and
  flybys, gives a stark contrast to the relatively quiescent history
  of h277.  At $z = 0$, h258 has a virial mass of $M_{vir} = 7.74
  \times 10^{11} \msun$, stellar mass $M_* = 4.46 \times 10^{10}
  \msun$, and maximum circular velocity $v_{\rm circ} = 242$ km s$^{-1}$.  %\editremark{Might we want to show the images of these galaxies?}

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{Figures/boring}
\includegraphics[width=\columnwidth]{Figures/exciting}
\caption{\label{fig:GalaxyImages}Synthetic SDSS $gri$ images of our two Milky Way-type galaxies, h277 (left) and h258 (right), created with \textsc{SUNRISE} \citep{Jonsson06}.
\label{fig:images}
}
\end{figure*}

@@

\begin{widetext}
% POINT:  Model
To characterize how much more likely LIGO will detect coalescing binaries with different masses, we use a common and
naive estimate for the volume to which advanced LIGO is sensitive \citep[see,e.g.,][]{PSellipticals}:\footnote{For
  simplicity, in this calculation we neglect the effects of cosmology; strong field coalescence; and black hole spin;
  see \cite{popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}, \cite{AstroPaper}, or \cite{RatesPaper} for more details.}
\begin{eqnarray}
V = \frac{4\pi}{3} [445 \unit{Mpc}]^3 \int p(m_1,m_2|Z) dm_1 dm_2 [ (\mc/1.2 M_\odot)^{5/6}]^3
\end{eqnarray}
where $\mc\equiv (m_1 m_2)^{3/5}/(m_1+m_2)^{1/5}$.   
This expression depends explicitly on an assumed and metallicity-dependent mass distribution for compact object
mergers, through the characteristic chirp mass $\mc_{*}(Z)\equiv [\int p(m_1,m_2|Z) dm_1 d,m_2 (\mc/1.2
  M_\odot)^{15/6}]^{5/16}$ .  We calibrate our metallicity-dependent mass distributions to metallicity-dependent binary
evolution calculations presented in \cite{popsyn-LowMetallicityImpact2-StarTrackRevised-2012} and
\cite{popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}.  For neutron
stars, we adopt a fiducial neutron star mass of $1.4 M_\odot$ at all metallicities.  For BH-NS binaries, we adopt a
highly simplified model: the black hole masses are uniformly  drawn from $5 M_\odot$ to the maximum 
black hole mass $M_{max}(Z)$ allowed by
\emph{isolated} stellar evolution, as reported in prior work \citep[see,e.g.][and references
  therein]{gwastro-EventPopsynPaper-2016}.  In agreement with much more detailed prior work \cite{popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}, this simplified model yields typical chirp masses for BH-NS binaries that
vary slightly as $Z/Z_\odot$ decreases, from a lower limit of $3 M_\odot$ near solar metallicity to an upper limit of
$4.3 M_\odot$ in low-metallicity environments.  Not least because the average chirp mass for BH-NS simply cannot vary
dramatically, given the functional form of $\mc$, our results do not depend sensitively on the choice of black hole mass
distribution.  Finally, for binary black holes, we assume  comparable-mass binaries
form ($m_1=m_2$), with the distribution of component masses chosen to be $\propto m_1^{-p}$ 
 between $5 M_\odot$ and $M_{max}(Z)$ and zero otherwise, adopting a fiducial exponent $p=2$; see, e.g.,
\cite{popsyn-LowMetallicityImpact2b-StarTrackRevised-2013}.    For binary black holes, this means  the
detection-weighted mass distribution [$\propto p(m_1,m_2)
\mc^{15/6}$] depends weakly on mass [$ p(m_1,m_2)
\mc^{15/6} \propto m_1^{-0.5} \delta(m_1-m_2)$.  In our simple model, the total  and chirp mass
distributions   are qualitatively consistent with detailed  binary evolution calculations \citep{popsyn-LowMetallicityImpact2b-StarTrackRevised-2013}.

@@

%\editremark{recheck numbers: we are tweaking things, and Zsun is seemingly inconsistent in the two approaches }
 Our two Milky Way-like galaxies have extremely similar star formation histories and metallicity evolution, particularly at late
 times.  
However, at early times, the \ExcitingGalaxy{} galaxy forms stars for $\simeq 2\unit{Gyr}$ at a lower characteristic metallicity
($Z\simeq 0.8 \times10^{-3}$) compared
to the \BoringGalaxy{} galaxy ($Z\simeq 2\times 10^{-3}$).  In this regime, the formation efficiency $\lambda$ and sensitive volume $V$
can depend sensitively on mass; for example, for binary black holes, the ratio of  $(\lambda V)_{\BoringGalaxy{}}/(\lambda
V)_{\ExcitingGalaxy{}} \simeq 3$.\footnote{Adopting a more extreme exponent for the metallicity dependence ($a=2$) only
  changes this ratio by of order $2$.}   However, in this same epoch, the star formation rate in the \ExcitingGalaxy{} (low-metallicity)
galaxy is smaller, by a factor of roughly 2.
% 1e-3:  V = 475, lambda= 20   
% 2e-3:  V= 318, lambda = 10
% so the ratio should be a factor of about 7x higher due to this region.
Therefore, because only a fraction of order $10\%$ \textbf{[checkme-JILLIAN?]} of all star formation occurs in this epoch, by this order of magnitude
argument, the overall number of
present-day coalescing binary black holes in our two galaxies is expected to differ by of order ten percent.
% 0.9 + 0.1*7/2
%
Using the concrete phenomenological calculations described above, we in fact find
$(r_{D}/M)_{\BoringGalaxy{}}/(r_{D}/M)_{\ExcitingGalaxy{}}\simeq 0.9$.

The close agreement between the two galaxies' binary black hole populations, determined by the anticorrelation between
star formation rate and metallicity, may be a single example of a broad trend.  This anticorrelation could cause
galaxies with similar present-day properties to always have similar present-day binary black hole populations,
regardless of their detailed assembly histories.
However, it is also possible that this close agreement is entirely a
a numerical coincdence between the specific star formation rate and metallicities in these two galaxies.  The effects described here
  are well within the scatter around the mass-metallicity relation.  Marginally different realizations of these
  histories can easily produce factors of order unity difference in galaxies with otherwise indistinguishable
  present-day properties.  That said, two other Milky-Way-like  galaxies (h239 and h285) also show very
  similar binary black hole coalescence rates $r_D/M$; see Table \ref{tab:Results} . \editremark{but: do more examples show dramatic differences, or comparable changes}

@@

\begin{table}
\begin{centering}
\begin{tabular}{llllc}\hline
\text{Simulation}  &  \text{BHBH} &  \text{BHNS}  & \text{NSNS} & M$_* (\msun)$\\
\hline
\hline
 \text{\BoringGalaxy} & 0.0224 & 0.000247 & 0.000206  & $4.24 \times 10^{10}$ \\
 \text{\ExcitingGalaxy}  &0.0216 & 0.000297 & 0.000228 & $4.46 \times 10^{10}$ \\
  \text{h239}  &  0.023 &X & 0.000268 & $4.50 \times 10^{10}$\\
 \text{h285} &  0.0236 & X & 0.000241 & $4.56 \times 10^{10}$ \\
 \text{\DwarfOne} &  0.0381 & 0.000308 & 0.000441& $7.8 \times 10^9$ \\
 \text{\DwarfTwo}  & 0.0949 & 0.000257 & 0.000884& $2.5 \times 10^8$ \\
\hline
\end{tabular}
\end{centering}
\caption{\label{tab:Results}\textbf{Event rates per unit mass $r_D/M$, arbitrary units}, for the three different types of binaries
  discussed, along with the stellar mass of each galaxy.  
}
\end{table}

>>>>

% POINT: Huge detail possible
The host galaxies of distant short GRBs have already been extensively investigated, with the associations being used to
draw preliminary conclusions about their progenitors  \citep{2014ARAA..52...43B}.   
%Of necessity, these analyses  on bulk correlations, often stacked to increase significance across multiple sources
%\editremark{clean up/don't piss off   Edo}.  
Unlike GRBs, detected gravitational wave sources will be limited by the range of LIGO to the local universe; for
example,  binary neutron star sources should be closer than $400\unit{Mpc}$.  Due to their proximity, each host galaxy
can be explored at great depth and detail via  position-resolved spectroscopy, enabling detailed position-resolved star-formation
and chemical evolution histories  \citep[see,e.g.][]{2009MNRAS.396..462K,2014MNRAS.444..336C,
CALIFA,CALIFA2}.  
%
However, unlike short GRBs and supernovae (SN), present-day compact  binary populations can depend sensitively on rare low metallicity
star formation.
% POINT: But is it useful?
In this work, we assess by concrete example the extent to which detailed analysis of individual galaxies' assembly histories will be essential in
investigations key physical questions about the origin of compact binary mergers.

This paper is organized as follows. 
In \S \ref{sec:sims}, we describe detailed hydrodynamical simulations of  several galaxies, including four of Milky Way-mass and two dwarfs.  Though the four Milky Way-like galaxies are morphologically similar at $z = 0$, we
show that their star formation and chemical evolution history have subtle but important differences due to their
distinctive merger histories.   To demonstrate
these differences have a practical impact,  in \S \ref{sec:model}, we introduce a simple,
metallicity-dependent  phenomenological model to calculate the present-day rate and mass distribution of compact binary
mergers from  a galaxy's known history.   In \S \ref{sec:results:BBH}, we use this model to investigate  the compact
binary  coalescence rate
dependence on the galaxy's assembly history.   
%% We also demonstrate that even the coalescence rates of binary neutron stars (NSNS) and black hole-neutron
%% star (BH-NS) binaries also depend surprisingly sensitively on the assembly history of each galaxy. 
In \S~\ref{sec:Discussion} we discuss the implications of our study for the interpretation of host galaxy associations
identified via transient multmiessenger astronomy, in the short and long term.  We summarize our results in \S~ \ref{sec:conclude}.
%% In \S \ref{sec:results} we show that while these two galaxy formation models should produce comparable detectable binary neutron star
%% populations, these two scenarios could produce dramatically different populations of electromagnetically detectable
%% merging black hole-neutron star binaries. 
%

@@

  The simulation h258, on the other hand, has a much
  more active merger history.  At $z = 1$ there is a 1:1 merger event;
  a gas disk rapidly reforms following the collision (see
  \citet{Governato09}), resulting in a massive disk galaxy at $z = 0$
  which looks remarkably similar to the Milky Way and to the other
  simulation, h277.  Prior to the $z = 1$ merger, each of the four
  progenitor galaxies actually experiences its own additional major
  merger events around $z = 3$.  The combination of
  the series of major mergers, plus a number of minor interactions and
  flybys, gives a stark contrast to the relatively quiescent history
  of h277.  At $z = 0$, h258 has a virial mass of $M_{vir} = 7.74
  \times 10^{11} \msun$, stellar mass $M_* = 4.46 \times 10^{10}
  \msun$, and maximum circular velocity $v_{\rm circ} = 242$ km s$^{-1}$.  %\editremark{Might we want to show the images of these galaxies?}

@@

\begin{figure*}
\includegraphics[width=\columnwidth]{Figures/boring}
\includegraphics[width=\columnwidth]{Figures/exciting}
\caption{\label{fig:GalaxyImages}Synthetic SDSS $gri$ images of two of our Milky Way-type galaxies, h277 (left) and h258 (right), created with \textsc{SUNRISE} \citep{Jonsson06}.
\label{fig:images}
}
\end{figure*}

@@

\begin{widetext}
% POINT:  Model
To characterize how much more likely LIGO will detect coalescing binaries with different masses, we use a common and
naive estimate for the volume to which advanced LIGO is sensitive \citep[see,e.g.,][]{PSellipticals}:\footnote{For
  simplicity, in this calculation we neglect the effects of cosmology; strong field coalescence; and black hole spin;
  see \cite{popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}, \cite{AstroPaper}, or \cite{RatesPaper} for more details.}
\begin{eqnarray}
V = \frac{4\pi}{3} [445 \unit{Mpc}]^3 \int p(m_1,m_2|Z) dm_1 dm_2 [ (\mc/1.2 M_\odot)^{5/6}]^3
\end{eqnarray}
where $\mc\equiv (m_1 m_2)^{3/5}/(m_1+m_2)^{1/5}$.   
This expression depends explicitly on an assumed and metallicity-dependent mass distribution for compact object
mergers, through the characteristic chirp mass $\mc_{*}(Z)\equiv [\int p(m_1,m_2|Z) dm_1 d,m_2 (\mc/1.2
  M_\odot)^{15/6}]^{5/16}$ .  We calibrate our metallicity-dependent mass distributions to metallicity-dependent binary
evolution calculations presented in \cite{popsyn-LowMetallicityImpact2-StarTrackRevised-2012} and
\cite{popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}.  For neutron
stars, we adopt a fiducial neutron star mass of $1.4 M_\odot$ at all metallicities.  For BH-NS binaries, we adopt a
highly simplified model: the black hole masses are uniformly  drawn from $5 M_\odot$ to the maximum 
black hole mass $M_{max}(Z)$ allowed by
\emph{isolated} stellar evolution, as reported in prior work \citep[see,e.g.][and references
  therein]{gwastro-EventPopsynPaper-2016}.  In agreement with much more detailed prior work \cite{popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}, this simplified model yields typical chirp masses for BH-NS binaries that
vary slightly as $Z/Z_\odot$ decreases, from a lower limit of $3 M_\odot$ near solar metallicity to an upper limit of
$4.3 M_\odot$ in low-metallicity environments.  Not least because the average chirp mass for BH-NS simply cannot vary
dramatically, given the functional form of $\mc$, our results for the BH-NS coalescence rate per unit galaxy mass do not depend sensitively on the choice of black hole mass
distribution.  Finally, for binary black holes, we assume  comparable-mass binaries
form ($m_1=m_2$), with the distribution of component masses chosen to be $\propto m_1^{-p}$ 
 between $5 M_\odot$ and $M_{max}(Z)$ and zero otherwise, adopting a fiducial exponent $p=2$; see, e.g.,
\cite{popsyn-LowMetallicityImpact2b-StarTrackRevised-2013}.    For binary black holes, this means  the
detection-weighted mass distribution [$\propto p(m_1,m_2)
\mc^{15/6}$] depends weakly on mass [$ p(m_1,m_2)
\mc^{15/6} \propto m_1^{-0.5} \delta(m_1-m_2)$.  In our simple model, the total  and chirp mass
distributions   are qualitatively consistent with detailed  binary evolution calculations \citep{popsyn-LowMetallicityImpact2b-StarTrackRevised-2013}.

@@

%\editremark{recheck numbers: we are tweaking things, and Zsun is seemingly inconsistent in the two approaches }
 Our two Milky Way-like galaxies have extremely similar star formation histories and metallicity evolution, particularly at late
 times.  
However, at early times, the \ExcitingGalaxy{} galaxy forms stars for $\simeq 2\unit{Gyr}$ at a lower characteristic metallicity
($Z\simeq 0.8 \times10^{-3}$) compared
to the \BoringGalaxy{} galaxy ($Z\simeq 2\times 10^{-3}$).  In this regime, the formation efficiency $\lambda$ and sensitive volume $V$
can depend sensitively on mass; for example, for binary black holes, the ratio of  $(\lambda V)_{\BoringGalaxy{}}/(\lambda
V)_{\ExcitingGalaxy{}} \simeq 3$.\footnote{Adopting a more extreme exponent for the metallicity dependence ($a=2$) only
  changes this ratio by of order $2$.}   However, in this same epoch, the star formation rate in the \ExcitingGalaxy{} (low-metallicity)
galaxy is smaller, by a factor of roughly 2.
% 1e-3:  V = 475, lambda= 20   
% 2e-3:  V= 318, lambda = 10
% so the ratio should be a factor of about 7x higher due to this region.
Therefore, because only a fraction of order $10\%$  of all star formation occurs in this epoch, by this order of magnitude
argument, the overall number of
present-day coalescing binary black holes in our two galaxies is expected to differ by of order ten percent.
% 0.9 + 0.1*7/2
%
Using the concrete phenomenological calculations described above, we in fact find
$(r_{D}/M)_{\BoringGalaxy{}}/(r_{D}/M)_{\ExcitingGalaxy{}}\simeq 0.9$.

The close agreement between the two galaxies' binary black hole populations, determined by the anticorrelation between
star formation rate and metallicity, may be a single example of a broad trend.  This anticorrelation could cause
galaxies with similar present-day properties to always have similar present-day binary black hole populations,
regardless of their detailed assembly histories.
However, it is also possible that this close agreement is entirely a
a numerical coincdence between the specific star formation rate and metallicities in these two galaxies.  The effects described here
  are well within the scatter around the mass-metallicity relation.  Marginally different realizations of these
  histories can easily produce factors of order unity difference in galaxies with otherwise indistinguishable
  present-day properties.  That said, two other Milky-Way-like  galaxies (h239 and h285) also show very
  similar binary black hole coalescence rates $r_D/M$; see Table \ref{tab:Results}. % \editremark{but: do more examples show dramatic differences, or comparable changes}

@@

\begin{table}
\begin{centering}
\begin{tabular}{llllc}\hline
\text{Simulation}  &  \text{BHBH} &  \text{BHNS}  & \text{NSNS} & M$_* (\msun)$\\
\hline
\hline
 \text{\BoringGalaxy} & 0.0224 & 0.000247 & 0.000206  & $4.24 \times 10^{10}$ \\
 \text{\ExcitingGalaxy}  &0.0216 & 0.000297 & 0.000228 & $4.46 \times 10^{10}$ \\
  \text{h239}  &  0.023 & 0.000354 & 0.000268 & $4.50 \times 10^{10}$\\
 \text{h285} &  0.0236 & 0.000284 & 0.000241 & $4.56 \times 10^{10}$ \\ \hline
 \text{\DwarfOne} &  0.0381 & 0.000308 & 0.000441& $7.8 \times 10^9$ \\
 \text{\DwarfTwo}  & 0.0949 & 0.000257 & 0.000884& $2.5 \times 10^8$ \\
\hline
\end{tabular}
\end{centering}
\caption{\label{tab:Results}\textbf{Event rates per unit mass $r_D/M$, arbitrary units}, for the three different types of binaries
  discussed, along with the stellar mass of each galaxy.  
}
\end{table}
